A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-21 paper.tex
Pass through latter sections

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practise in coalescent theory.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence,
in terms of the level of detail inferred about recombination,
lie somewhere
in between a sequence of unrelated trees on one extreme and a
completely precise event ARG on the other.

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context. The
trend to decouple the ARG from its origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
substantially \emph{larger} than the usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with large samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may help improve
methods for inference method evaluation. In the majority of cases,
ARG inference accuracy is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree metrics as originally proposed by \citet{kuhner2015assessing}.
While this provides valuable insights into the method's performance,
it is not clear how well these tree distance metrics reflect
performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in a very indirect manner.
Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
applied to the very large sample sizes currently of interest.
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which is a useful addition but only indirectly captures
topological comparisons.
% Not sure this is the right place for this sentence
All of these comparisons and evaluations have been based on
highly idealised simulations, and the effects richness of real data
(beyond simplistic error models) are almost entirely unknown.
A suite of true ARG distance metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

>>>>

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer complete eARGs (section XXX),
which contain specific causal recombination \emph{events}
for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer causal events but rather
infer gARGs with differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
Therefore, they contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.

A necessary first step towards this goal is some terminological
clarity. As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
meaning that sample size is substantially larger than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may help
to clarify inference goals and improve
methods for evaluation.
In the majority of cases,
ARG inference accuracy is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
as originally proposed by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in a indirect manner, and the tree
metrics assume unlabelled nodes, effectively discarding information
about node sharing between trees.
Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance
based approaches, and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which captures many important aspects of dated ARGs
but neglects [SOME OTHER THINGS].
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes, and compared this to the
outputs of several methods.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
% Not sure this is the right place for this sentence
All of these evaluations have been based on
ground truth data from highly idealised simulations
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

[NOTE This is a pedantic and slightly bullshitty to end up on,
just writing it down to see if it works. Should probably cut.]
A final point to note, is that
if we are using the term ARG
to loosely refer to genetic genealogies of varying
levels of detail, either inferred or simulated,
it makes little sense to refer ``the'' Ancestral Recombination
Graph (i.e, in  the singular, like ``the tree'' or
``the phylogeny'')
% TODO add more if we keep this
as favoured by some~\citep[e.g.][]{
arenas2013importance,hubisz2020inference,zhang2023biobank}.
While there is of course a single ``true'' ARG that describes
the actual history of a sample, in reality there will be astronically
many ARGs that are compatible with the data, even under
parsimony conditions or when sampling from a model.
There are many different levels of granularity at which we
can reason about recombination, and many different ARGs.
We should embrace this diversity.
